Eliminating prior-bias from sparse-projection tomographic reconstructions

### 2020-07-10 paper/tci.tex
minor changes to eq 1 and related work section

However, in the last decade, reconstruction from reduced views has
been made possible by assuming the data to exhibit sparsity under
certain mathematical transforms $\boldsymbol{\Upsilon}$ such as the
wavelet transforms, or the Discrete Cosine Transform (DCT). This is
known as sparsity prior and is the fundamental principle in the widely
used Compressed Sensing (CS) technique~\cite{Donoho,introCS}.  There
are multiple ways to incorporate the sparsity prior using CS. We use
the LASSO (least absolute shrinkage and selection operator) which
iteratively solves for the solution by penalizing a combination of
least squares error and $L_1$-norm of the sparse coefficients
$\boldsymbol{\theta}$ of the object $\boldsymbol{x}$. If $\boldsymbol{\Psi}$
represents the sparsity basis i.e. if $\boldsymbol{x} = \boldsymbol{\Psi\theta}$, and 
$\boldsymbol{\mathcal{R}}$ represents the acquisition model, then the
LASSO solution is described by one that minimizes
 \begin{equation}
J_{\text{CS}}(\boldsymbol{\theta}) = \lVert\boldsymbol{\mathcal{R} x}- \boldsymbol{y}\rVert_2^2  + \lambda_1\lVert\boldsymbol{\theta}\rVert_1
\label{Eq:simple_cs}
 \end{equation}
 We solve this cost function using the popular l1-regularized least
 squares (l1-ls) package available in~\cite{l1ls}, and choose DCT as
 our sparsity basis $\boldsymbol{\Upsilon}$.  Fig.~\ref{fig:story}
 (second column) demonstrates the benefit of using sparsity prior when
 the number of projection views is limited (100 views).

@@

 This paper is organized as follows. After discussing related work in
 Section~\ref{sec:related}, Section~\ref{sec:contributions} lays down
 the key contributions of this work. Section~\ref{sec:tmh}
 demonstrates the utility of both the uniform and spatially-varying
 methods on a longitudinal medical dataset. We now move to the
 details. Section~\ref{sec:method_uniform_prior} describes the
 construction of the uniform eigenspace prior, followed by the
 corresponding results in real and synthetic 3D biological datasets in
 Section~\ref{sec:results_uniform_prior}. Section~\ref{sec:method_spatially_varying_prior}
 describes how the uniform prior needs to be modified when accurate
 details of new changes are to be observed. A spatially-varying
 technique offers a solution, the results of which are presented in
 Section~\ref{sec:results_spatially_varying_prior}. Section\ref{sec:discussion}
 discusses tuning of the hyperparameters involved, and limitations of
 our method. Finally, we conclude with key inferences that can be
 drawn from our work in Section~\ref{sec:conclusions}.

@@

 \section{Related Work}
\label{sec:related}
There exist techniques that aim to reconstruct from few-views in
specialized applications. For example, in~\cite{Essam2015},
sparsity-constrained optimization is presented for angiography. Here,
the regions of interest are the vessels alone and they are highlighted
by physically inserting a contrast agent, and therefore there is a
inherent sharper contrast between the vessels and the
background. In many other general application scenarios like the ones presented in this paper,
any information about the regions of interest is not known. In other applications where the spatial-gradient of the underlying volume is known to be sparse, the `total variation' method, as used in~\cite{Li2015,Polak2017} can be applied. As we will later see in Figs.\ref{fig:object-prior_test_okra} and~\ref{fig:object-prior_test_sprouts}, the spatial-gradient of many datasets is not sparse.

Most object-prior based reconstruction has been in the areas of
dynamic CT and 4D-CT. In dynamic CT, the object being scanned
undergoes changes between successive projection views and hence is not
stationary even within a single complete scan. In contrast, in 4D-CT,
a set of 3D CT scans are taken over short intervals of time and all the 3D volumes are reconstructed together. Here, the object undergoes changes between successive scans. ~\cite{PICCS} is one of the earliest work to establish the use of object-prior along with sparsity prior in dynamic CT. Its limitation was the unchecked over-emphasis of the prior on the reconstruction of the test at hand. This limitation was later dealt with by applying very specific object-properties for
reconstruction. For example, in~\cite{Van2015}, the knowledge of the
attenuation coefficient of the fluid is used as a prior for
reconstructing its flow through gravel. In~\cite{Koen2020}, the
changes across the sucessive scan volumes is assumed to be continuous,
thus enabling the use of optical flow to model the motion between
corresponding voxels of different scan volumes. In another
study~\cite{vincent2017}, changes between successive scans were
modelled by affine deformation whose parameters were computed for
motion-correction. Such a correction enables acquisition from fewer
views than would have otherwise not been possible for reconstruction
with adequate fidelity. In~\cite{daniil2015}, all the scan volumes are
reconstructed together using spatio-temporal regularization.

\section{Contributions}
\label{sec:contributions}
This paper focusses on few-views reconstruction with an emphasis on
longitudinal studies.  Fig.~\ref{fig:prior_overview} provides an
overview. In contrast to the object prior-based studies
mentioned above, we reconstruct the current test object without any
assumption of continuity of changes or some knowledge of the
attenuation coefficients of the structures. We do not make any
temporal assumption in terms of time intervals such as in 4D CT
scans. We use the current measurements from few-views and previous
scans of the same object. A key idea in our work is the starting point
-- the new test volume is close to the space spanned by the
eigenvectors of the multiple representative previously scanned
objects.

@@

>>>>

However, in the last decade, reconstruction from reduced views has
been made possible by assuming the data to exhibit sparsity under
certain mathematical transforms $\boldsymbol{\Upsilon}$ such as the
wavelet transforms, or the Discrete Cosine Transform (DCT). This is
known as sparsity prior and is the fundamental principle in the widely
used Compressed Sensing (CS) technique~\cite{Donoho,introCS}.  There
are multiple ways to incorporate the sparsity prior using CS. We use
the LASSO (least absolute shrinkage and selection operator) which
iteratively solves for the solution by penalizing a combination of
least squares error and $L_1$-norm of the sparse coefficients
$\boldsymbol{\theta}$ of the object $\boldsymbol{x}$. If $\boldsymbol{\Psi}$
represents the sparsity basis, i.e., if $\boldsymbol{x} = \boldsymbol{\Psi\theta}$, and 
$\boldsymbol{\mathcal{R}}$ represents the acquisition model, then the
LASSO solution is described by one that minimizes
 \begin{equation}
J_{\text{CS}}(\boldsymbol{\theta}) = \lVert\boldsymbol{\mathcal{R} \Psi\theta}- \boldsymbol{y}\rVert_2^2  + \lambda_1\lVert\boldsymbol{\theta}\rVert_1
\label{Eq:simple_cs}
 \end{equation}
 We solve this cost function using the popular l1-regularized least
 squares (l1-ls) package available in~\cite{l1ls}, and choose DCT as
 our sparsity basis $\boldsymbol{\Upsilon}$.  Fig.~\ref{fig:story}
 (second column) demonstrates the benefit of using sparsity prior when
 the number of projection views is limited (100 views).

@@

 This paper is organized as follows. After discussing related work in
 Section~\ref{sec:related}, Section~\ref{sec:contributions} lays down
 the key contributions of this work. Section~\ref{sec:tmh}
 demonstrates the utility of both the uniform and spatially-varying
 methods on a longitudinal medical dataset. We now move to the
 details. Section~\ref{sec:method_uniform_prior} describes the
 construction of the uniform eigenspace prior, followed by the
 corresponding results in real and synthetic 3D biological datasets in
 Section~\ref{sec:results_uniform_prior}. Section~\ref{sec:method_spatially_varying_prior}
 describes how the uniform prior needs to be modified when accurate
 details of new changes are to be observed. A spatially-varying
 technique offers a solution, the results of which are presented in
 Section~\ref{sec:results_spatially_varying_prior}. Section~\ref{sec:discussion}
 discusses tuning of the hyperparameters involved, and limitations of
 our method. Finally, we conclude with key inferences that can be
 drawn from our work in Section~\ref{sec:conclusions}.

@@

 \section{Related Work}
\label{sec:related}
The idea of using a reduced number of views is most pronounced in
specialized applications. For example, in~\cite{Essam2015},
sparsity-constrained optimization is presented for angiography. Here,
the regions of interest are the vessels alone and they are highlighted
by physically inserting a contrast agent, and therefore there is a
inherent sharper contrast between the vessels and the background.  In
other applications where the spatial-gradient of the underlying volume
is known to be sparse, the `total variation' method, as used
in~\cite{Li2015,Polak2017} can be applied.  In a more generalized
application scenario, like the ones addressed in this paper, any
information about the regions of interest is not known, and such
techniques have limited applicability.

%  As can be seen
% in the situatiowe will
% later see in Figs.\ref{fig:object-prior_test_okra}
% and~\ref{fig:object-prior_test_sprouts}, the spatial-gradient of many
% datasets is not sparse.

Most object-prior based reconstruction has been in the areas of
dynamic CT and 4D-CT. In dynamic CT, the object being scanned
undergoes changes between successive projection views and hence is not
stationary even within a single complete scan. In contrast, in 4D-CT,
a set of 3D CT scans are taken over short intervals of time, and the
object undergoes changes between successive scans.  The reconstruction
is then performed over the entire set generating several 3D
volumes. ~\cite{PICCS} is one of the earliest work to establish the
use of object-prior along with sparsity prior in dynamic CT, in an
attempt to reduce the number of views. The limitation is the
unchecked over-emphasis of the prior on the reconstruction of the test
at hand. This limitation was later dealt with by applying very
specific object-properties for reconstruction. For example,
in~\cite{Van2015}, the knowledge of the attenuation coefficient of the
fluid is used as a prior for reconstructing its flow through
gravel. In~\cite{Koen2020}, the changes across the sucessive scan
volumes is assumed to be continuous, thus enabling the use of optical
flow to model the motion between corresponding voxels of different
scan volumes. In another study~\cite{vincent2017}, changes between
successive scans were modelled by affine deformation whose parameters
were computed for motion-correction. Such a correction enables
acquisition from fewer views than would have otherwise not been
possible for reconstruction with adequate
fidelity. In~\cite{daniil2015}, all the scan volumes are reconstructed
together using spatio-temporal regularization.

\section{Contributions}
\label{sec:contributions}
This paper focusses on few-views reconstruction with an emphasis on
longitudinal studies. In contrast to the object prior-based studies mentioned
above, we reconstruct the current test object without any assumption
of continuity of changes or some knowledge of the attenuation
coefficients of the structures. We do not make any temporal assumption
in terms of time intervals -- prior scans could be months apart. We
use the current measurements from few-views and previous scans of the
same object. A key idea in our work is the starting point -- the new
test volume is close to the space spanned by the eigenvectors of the
multiple representative previously scanned objects.

@@

Fig.~\ref{fig:prior_overview} provides an overview.
